---
title: "P8105 Homework 2"
author: "Diana Hernandez"
date: "2023-10-04"
output: github_document
---

# load tidyverse

```{r}
library(tidyverse)
```

# Problem 1: using FiveThirtyEight dataset

# Data wrangling of pols_month_data

```{r}
pols_month_data = 
  read_csv("./fivethirtyeight_datasets/pols-month.csv") |>
  separate(col=mon, into=c('year', 'month', 'day'), sep='-') |>
  mutate(
    month = case_match(
      month,
      "01" ~ "January",
      "02" ~ "February",
      "03" ~ "March",
      "04" ~ "April",
      "05" ~ "May",
      "06" ~ "June",
      "07" ~ "July",
      "08" ~ "August",
      "09" ~ "September",
      "10" ~ "October",
      "11" ~ "November",
      "12" ~ "December"
    ),
    president =
      if_else(prez_gop==1, paste("Republican"), paste("Democrat"))
              )|>
  select(-prez_dem, -prez_gop, -day)
```

# Data wrangling of snp_data

```{r}
snp_data = 
  read_csv("./fivethirtyeight_datasets/snp.csv") |>
  separate(col=date, into=c('month', 'day', 'year'), sep='/') |>
  select('year', 'month', 'day', everything()) |>
  mutate(
    month = case_match(
      month,
      "1" ~ "January",
      "2" ~ "February",
      "3" ~ "March",
      "4" ~ "April",
      "5" ~ "May",
      "6" ~ "June",
      "7" ~ "July",
      "8" ~ "August",
      "9" ~ "September",
      "10" ~ "October",
      "11" ~ "November",
      "12" ~ "December"),
    year = if_else(year < 50, paste('20',year), paste('19',year)),
    year = gsub(" ","",year) 
    ) |>
  select(-day)
```

# Data wrangling of unemployment_data

```{r}
unemployment_data = 
  read_csv("./fivethirtyeight_datasets/unemployment.csv") |> 
  janitor::clean_names() |>
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment"
  ) |> 
  mutate(
    month = case_match(
      month,
      "jan" ~ "January",
      "feb" ~ "February",
      "mar" ~ "March",
      "apr" ~ "April",
      "may" ~ "May",
      "jun" ~ "June",
      "jul" ~ "July",
      "aug" ~ "August",
      "sep" ~ "September",
      "oct" ~ "October",
      "nov" ~ "November",
      "dec" ~ "December"),
    year = as.character(year)
    )
```

# Data merging of the FiveThirtyEight datasets

```{r}
five_thirty_eight_data =
  inner_join(pols_month_data, snp_data)

final_five_thirty_eight_data =
  inner_join(five_thirty_eight_data, unemployment_data)
```

The pols_month dataset includes the following variables: year, month, gov_gop, sen_gop, rep_gop, gov_dem, sen_dem, rep dem, and president. The snp dataset includes the following variables: year, month, and close. The unemployment dataset includes the following variables: year, month, and unemployment rate.

The final FiveThirtyEight dataset includes the following datasets: pols_month, snp, and unemployment. The final dataset has 786 rows and 11 columns. There are 11 key variables: year, month, gov_gop, sen_gop, rep_gop, gov_dem, sen_dem, rep_dem, president, close, and unemployment. The years range from 1950 to 2015. 

# Problem 2: using Mr. Trash Wheel dataset

# Load readxl to import Excel dataset

```{r}
library(readxl)
```

# Data wrangling of Mr. Trash Wheel dataset

```{r}
mr_trash_wheel_data = 
  read_excel("./trashwheel_dataset/trash_wheel_collection_data.xlsx", 
             n_max = 584) |>
  janitor::clean_names() |>
  select(-homes_powered, -15, -16) |>
  mutate(
    homes_powered = (500*weight_tons)/30,
    dataset = "Mr. Trash Wheel",
    year = as.numeric(year)
    )
```

# Data wrangling of Professor Trash Wheel dataset 

```{r}
professor_trash_wheel_data =
  read_excel("./trashwheel_dataset/trash_wheel_collection_data.xlsx", 
             2, n_max = 106) |>
  janitor::clean_names() |>
  select(-homes_powered) |>
  mutate(
    homes_powered = (500*weight_tons)/30,
    dataset = "Professor Trash Wheel")
```

# Data wrangling of Gwynnda dataset

```{r}
gwynnda_trash_wheel_data =
  read_excel("./trashwheel_dataset/trash_wheel_collection_data.xlsx", 
             4, n_max = 155) |>
  janitor::clean_names() |>
  select(-homes_powered) |>
  mutate(
    homes_powered = (500*weight_tons)/30,
    dataset = "Gwynnda Trash Wheel")
```

# Data merging of the Trash Wheel datasets

```{r}
final_trash_wheel_data = 
  bind_rows(mr_trash_wheel_data, professor_trash_wheel_data, gwynnda_trash_wheel_data)

filter(gwynnda_trash_wheel_data, month=="July", year=="2021") 
```

The total number of observations in the final data set called `final_trash_wheel_data` is `r length(final_trash_wheel_data$dumpster)`. Examples of key variables in `final_trash_wheel_data` are: dumpster number, month, year, date, weight in tons, homes powered, name of data set (Mr.Trash Wheel, Professor Trash Wheel, and Gwynnda Trash Wheel). I calculated `homes_powered` for each individual data set prior to merging each data set using `bind_rows`.

The total weight of trash collected in the Professor Trash Wheel dataset is `r sum(professor_trash_wheel_data$weight_tons)` tons.

In July 2021, the total number of cigarette butts collected by Gwynnda was `r sum(gwynnda_trash_wheel_data$cigarette_butts)` cigarette butts.

# Problem 3: using data_mci

# Data wrangling of mci_amyloid dataset

```{r}
mci_amyloid_data = 
  read_csv("./data_mci/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names() |>
  pivot_longer(
    baseline:time_8,
    names_to = "visit",
    names_prefix = "time_",
    values_to = "time") |>
  mutate(
    visit = replace(visit, visit == "baseline", "0")
  )
```

In the import process,  I skipped the first row to have cases start at row 1 using `skip`. I also used `janitor::cleannames()` to make the variable names tidy. I used `pivot_longer` to not have each visit spread across 5 columns. I also changed the visit named `baseline` to `0` in order to have a more uniform list of visit names using `replace`. I deleted "time_" in front of each visit as it was repetitive using `names_prefix `. `mci_amyloid_data` consists of study IDs, visits 0, 2, 4, 6, and 8, and time (in years) elapsed since the study baseline to the visit where biomarker Amyloid was measured.

# Data wrangling of mci_baseline dataset 

```{r}
mci_baseline_data = 
  read_csv("./data_mci/MCI_baseline.csv", skip = 1, na = ".") |>
  janitor::clean_names() |>
  mutate(
    sex =
      case_match(
        sex,
        0 ~ "Female",
        1 ~ "Male"),
    sex = as.factor(sex),
    apoe4 =
      case_match(
        apoe4,
        0 ~ "APOE4 non-carrier",
        1 ~ "APOE4 carrier"),
    apoe4 = as.factor(apoe4)
    ) |>
  rename(age_at_baseline = current_age,
         study_id = id) |>
  filter(age_at_baseline < age_at_onset | is.na(age_at_onset))
 

mci_in_study_data =
    read_csv("./data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |>
  mutate(
    sex =
      case_match(
        sex,
        0 ~ "Female",
        1 ~ "Male"),
    sex = as.factor(sex),
    apoe4 =
      case_match(
        apoe4,
        0 ~ "APOE4 non-carrier",
        1 ~ "APOE4 carrier"),
    apoe4 = as.factor(apoe4)
    ) |>
  rename(age_at_baseline = current_age,
         study_id = id) |>
  filter(age_at_baseline < age_at_onset) 

filter(mci_baseline_data, sex=="Female", apoe4=="APOE4 carrier")
```

In the import process, I skipped the first row to have cases start at row 1 using `skip`. I also used `janitor::cleannames()` to make the variable names tidy. I changed `current_age` to `age_at_baseline` to follow a similar pattern to `age_at_onset`. I changed `id` to `study_id` to match the `mci_amyloid_data` data set. The data sets `mci_baseline_data` and `mci_in_study` include the following variables: study ID, sex (male or female), education level, age at baseline of study, age at onset of MCI, and APOE4 carrier status. I created one data frame called `mci_baseline_data` with every eligible participant along with missing data. In the `mci_baseline_data`, I used `filter` to only keep eligible participants (participants whose age at onset was older than their age at baseline). I also kept missing data in `mci_baseline_data`. I created another data frame called `mci_in_study_data` with participants who developed MCI during the study using a `filter`. I made sex and APOE carrier status not numeric using `case_match`. Sex was labeled as female or male. APOE carrier status was labeled as APOE carrier or APOE non-carrier. 

The study recruited `r length(mci_baseline_data$study_id)` eligible participants. Only `r length(mci_in_study_data$age_at_onset)` participants developed MCI during the study.

The mean age at baseline is `r mean(mci_baseline_data$age_at_baseline)` years old. 

There are `r length(mci_in_study_data$study_id)` females that are APOE4 carriers in the study.

# Merging mci datasets

```{r}

differences_baseline = 
  anti_join(mci_baseline_data, mci_amyloid_data, by = "study_id")

final_mci_data = 
  inner_join(mci_baseline_data, mci_amyloid_data)
```

I used `anti_join` to see if there were some participants appear in only the baseline or amyloid datasets. `inner_join` was used to only keep participants who appear in both data sets since I noticed that not every participant was present in both data sets. The resulting dataset (`final_mci_data`) includes the following variables: study ID, age at baseline, sex (male or female), education, APOE4 carrier status, age at onset of MCI, visit number (0, 2, 4, 6, 8), and time (in years) elapsed since the study baseline to the visit where biomarker Amyloid (42/40 ratio) was measured.
